---
title: "Faster simulation using revdbayes and Rcpp"
author: "Paul J. Northrop"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Faster simulation using revdbayes and Rcpp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: revdbayes.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

This vignette introduces a new feature of *revdbayes*: reducing posterior simulation time by performing the most time-consuming tasks using C++ functions.  This achieved using a new facility in the *rust* package [@rust], which in turn uses the **Rcpp** package [@Rcpp, @RcppDEbook]. The result is a new function `rpost_rcpp`, which has the same structure as the existing function `rpost`.  From a user's perspective the only difference between these two functions occurs if they wish to supply their own prior distribution: `rpost_rcpp` requires an external pointer to a C++ function (see [Providing a user-defined prior](#cpp_fun)), whereas `rpost` requires input R functions (see the vignette
[Introducing revdbayes](revdbayes-vignette.html)).

Before we deal with user-supplied priors we compare posterior simulation times using `rpost` and `rpost_rcpp` for examples based on in-built prior distributions.  We use the default settings of `rpost` and `rpost_rcpp` throughout.  We also compare the speed of these functions with the function `posterior` in the **evdbayes** package [@evdbayes].

## Performance comparisons

```{r}
library(revdbayes)
library(evdbayes)
library(microbenchmark)
# Set the number of posterior samples required.
n <- 1000
set.seed(46)
```

### Generalised Pareto (GP) model

Use repeat the analysis of the Gulf of Mexico Wave Height Data from the [Introducing revdbayes](revdbayes-vignette.html)) vignette, using the **microbenchmark** package [@microbenchmark] to check that using Rcpp does indeed reduced computation time.

```{r}
data(gom)
u <- quantile(gom, probs = 0.65)
fp <- set_prior(prior = "flat", model = "gp", min_xi = -1)
res <- microbenchmark(
  rpost = rpost(n = n, model = "gp", prior = fp, thresh = u, data = gom),
  rpost_rcpp = rpost_rcpp(n = n, model = "gp", prior = fp, thresh = u, data = gom)
)
print(res, signif = 3)
options(microbenchmark.unit = "relative")
print(res, signif = 2)
```

In this example `rpost_rcpp` is, on average, a factor of approximately 7-8 times faster than `rpost`.

### The Generalised Extreme Value (GEV) model

Use repeat the analysis of the Port Pirie annual maximum sea level data from the [Introducing revdbayes](revdbayes-vignette.html)).  We add to the comparison of performances the example calculations that feature in the **evdbayes** user guide based on the `posterior` function.  

```{r}
data(portpirie)
mat <- diag(c(10000, 10000, 100))
pn <- set_prior(prior = "norm", model = "gev", mean = c(0,0,0), cov = mat)
# Tuning parameters from the evdbayes user guide.
t0 <- c(3.87, 0.2, -0.05) 
s <- c(.06, .25, .25)
b <- 0
res <- microbenchmark(
  rpost = rpost(n = n, model = "gev", prior = pn, data = portpirie),
  rpost_rcpp = rpost_rcpp(n = n, model = "gev", prior = pn, data = portpirie),
  evdbayes = posterior(n + b - 1, t0, prior = pn, lh = "gev", data = portpirie,
                       psd = s, burn = b)
)
options(microbenchmark.unit = NULL)
print(res, signif = 3)
options(microbenchmark.unit = "relative")
print(res, signif = 2)
```

The comparison above is generous to `posterior` because the burn-in has been set to zero and `posterior` produces a dependent sample rather than a random sample.  The *effective sample size* of an MCMC sample from `posterior` varies between simulations and across parameters.  Repeated use of the `effectiveSize` function in the **coda** package [@coda] suggests that the effective sample size in this example is of the order of 100 to 200, whereas the **revdbayes** functions `rpost` and `rpost_rcpp` produce random samples of size 1000. `rpost` is less than 50\% slower than `posterior` while `rpost_rcpp` is a factor of approximately 4 faster than `posterior`.

## Providing a user-defined prior {#cpp_fun}

## References

